### 2025-03-25 paper/main.tex
CH content; fourier triangle area

% bibliography issues
\usepackage{natbib}
\bibliographystyle{aasjournal}
\setcitestyle{aysep={}}
\setlength{\bibsep}{0pt plus 0.3ex}

@@

\medskip
\paragraph{Abstract}
Any permutation-invariant function of data points $\vec{r}_i$
can be written in the form $\rho(\sum_i\phi(\vec{r}_i))$,
where $\rho$ and $\phi$ are functions.
This form---known in the machine-learning literature as Deep Sets---also generates a map--reduce algorithm.
The area of a triangle is a permutation-invariant function of the locations $\vec{r}_i$ of the three corners $1\leq i\leq 3$.
We find the formula for the area of a triangle that is explicitly in Deep Sets form.
This project was motivated by questions about the fundamental computational complexity of $n$-point statistics in cosmology; that said, no insights of any kind were gained from these results.

@@

A function $f(\vec{r}_1, \dots, \vec{r}_N)$ is invariant to permutation of its arguments if and only if it can be decomposed in the form
\begin{equation}
    f(r_1, \dots, r_N) = \rho \, \big( \sum_{i=1}^{N} \, \phi(\vec{r}_i) \big)~,\label{eq:DeepSets}
\end{equation}
for some functions $\rho$ and $\phi$. This result---often known by the name of the deep learning architecture derived from it, \emph{Deep Sets} (\citealt{Zaheer+17deepsets}; see Theorem~2)---is not at all obvious.
For example, even though Heron's formula \eqref{eq:Heron} is explicitly permutation-invariant, it is not actually in the form of \eqref{eq:DeepSets}. 
Further, despite the triangle area being permutation-invariant, there is no known (prior to this work) formula for the area of a triangle that is explicitly in Deep Sets form.

While this result was recently popularized as Deep Sets, it follows from the classical theory of invariant polynomials.
Specifically, it is known that the permutation-invariant polynomials on a set of any-dimensional data are representable in terms of the polarizations of the permutation-invariant polynomials on a corresponding set of one-dimensional data \citep{Weyl1939}.
Further, the permutation-invariant polynomials on a set of one-dimensional data are representable in terms of the power sums $\sum_{i=1}^{n} r_i^j$ \citep{Waring1782}.
This matches onto the Deep Sets form, as it states that a permutation-invariant polynomial of any-dimensional data $f(\{r_i\})$ can be decomposed into a polynomial ($\rho$) of polarizations of the power sums ($\sum_i \phi(\vec{r}_i)$).
\CH{this is mostly plagiarized from BBS email; need to understand it better to massage it.} \Hogg{Important to emphasize that there is a proof of universality, and it's old.}

@@

Further, and important to our own motivations, cosmological clustering statistics such as 
the correlation function \citep{Peebles1973},
the power spectrum \citep{Peebles1973},
the three-point correlation function \citep{PeeblesGroth1975},
the bispectrum \citep{FrySeldner1982},
and higher-order statistics \citep{Peebles1980book}
used for precision measurement in cosmology \citep[e.g.,][]{Planck18CosmoParams,Planck18PNG,Planck18Inflation,Cabass+2022}
are all permutation-invariant.
Calculating these statistics na\"ively involves loops over all $k$-tuplets of points, which is not only inconsistent with Deep Sets form, but also has a runtime that scales as $N^k$, where $N$ the number of points in the set.

@@

@@

\secbreak
We now consider two more applications of the Deep Sets form to Heron's formula and to the potential energy of a system of point masses.
We stated above that Heron's formula \eqref{eq:Heron} is explicitly permutation-invariant, and yet it is not represented in Deep Sets form. Consider instead
\begin{equation}
\label{eq:HeronDeepSets}
    \Delta = \frac{1}{4} \, \sqrt{
        \left( a^2 + b^2 + c^2 \right)^2
        - 2 \left( a^4 + b^4 + c^4 \right)
    }~.
\end{equation}
It is simple to show that this is exactly equivalent to Heron's formula when expanded, but it is now in Deep Sets form.\footnote{
    This is not novel. Heron's formula is a special case of Brahmagupta's formula, which gives the area $K$ of a convex cyclic quadrilateral in terms of its four side lengths $a, b, c, d$ \citep{coolidge1939}.
    On Wikipedia, it has been written
    \begin{equation}
        K = \frac{1}{4} \, \sqrt{
            \left( a^2 + b^2 + c^2 + d^2 \right)^2
            + 8 \, a\,b\,c\,d
            - 2 \left( a^4 + b^4 + c^4 + d^4 \right)
        }~,
    \end{equation}
    which is exactly \eqref{eq:HeronDeepSets} if $d$ is zero.
    This expression is in Deep Sets form, with the product $a \, b \, c \, d$ 
    representable as $\exp( \sum_{x \in \{ a,b,c,d \}} \log x )$.
}

\CH{feels out of place...} \Hogg{This can be deleted.}
The area of a triangle is not only invariant to permutation, but also to translation, rotation, and reflection.
Formulae \eqref{eq:sine} and \eqref{eq:Heron}, by virtue of referencing only side-lengths and angles, are explicitly invariant to these other symmetries.
While true, it is not obvious that our solution \eqref{eq:result} is so invariant.
An explicit proof of this is left to the reader.

While the triangle area is related to physics problems, the polynomial methods we employ to find the Deep Sets form for the triangle area \eqref{eq:result} are not relevant in a physical context.
More relevant instead is to consider a $\phi$ which Fourier transforms the input data.
In doing so, translation symmetry is encoded as well, since the Fourier basis $\exp (i\, \vec{k}\cdot\vec{x})$ is the eigenbasis of the translation operator.
Below, we show that this can be used to produce a Deep Sets form for the gravitational potential energy of a system of point masses.

Consider a system of $N$ point particles with masses $m_i$ and positions $\vec{r}_i$.
The total energy is given in Deep Sets form by
\begin{align}
    \phi_{\vec{k}}(m_j, \vec{r}_j)
    &=m_j \, e^{-i \, \vec{k} \cdot \vec{r}_j}
    ~,
    \quad
    \sum_{i=1}^{N} \phi_{\vec{k}}(m_i, \vec{r}_i)
    = \mathcal{F}[\rho](\vec{k})~,
    \nonumber\\
    U \, \big( \sum_{i=1}^{N} \, \phi_{\vec{k}}(m_i, \vec{r}_i) \big)
    &= -\frac{G}{4 \pi^2} \,
        \int \d^3 r \,
        \mathcal{F}^{-1} \left[
            \mathcal{F}[\rho] \ast \left(
            k^{-2} \, \mathcal{F}[\rho] \right)
        \right]~.
    \label{eq:EnergyDeepSets}
\end{align}
A derivation is given in the \nameref{sec:appendix}.
In principle this requires $\phi$ to be evaluated at every possible wavenumber $\vec{k}$.
To avoid making this infinite-dimensional $\phi$, one can instead evaluate $\mathcal{F}[\rho]$ at a finite set of wavenumbers $\vec{k}$ and live with approximations to the integrals necessary for the convolution and inverse Fourier transform.
The approximation will improve as the size of $\phi$ grows.
Approaches akin to this Fourier approach can be used \citep[and are used; e.g.,][]{Portillo+2018,Philcox+2022ENCORE} to estimate or approximate $k$-point statistics in cosmology, which was our motivating question.

We leave the reader with the following questions. 
It is trivial to take the form of \eqref{eq:result} and evaluate the sums for more than three vectors; does the result have any meaning?
Is there a generalization of the triangle area \eqref{eq:result} and simplex volume \eqref{eq:simplex} results to a $d$-simplex?
Our formulae \eqref{eq:result} and \eqref{eq:simplex} only used invariant factors of degree $2$, with exponents that are larger than $2$. Does this generalize to higher dimension simplexes, and can it be shown that higher order factors aren't needed?
One can replace the sum in \eqref{eq:DeepSets} with any other permutation-invariant aggregator (e.g., max or product), though we do not know if there are any claims to universality in these cases. Does another area formula exist for these other aggregators?
We have discussed the general value of a map function $\phi$ which Fourier transforms the points;\footnote{%
\Hogg{The Deep Sets version of the Fourier transform is reminiscent of a positional encoding (CITE), similar to those used to preserve ordering information in (otherwise permutation invariant) transformers (CITE).}}
does a formula like \eqref{eq:EnergyDeepSets} exist for the area of a triangle?

{\footnotesize\par\bigskip\noindent
This project would not have been possible without originating inspiration and discussions with Soledad Villar (JHU).
The authors thank Ben Blum-Smith (JHU), Jonathan Goodman (NYU), Wilson Gregory (JHU), Teresa Huang (Flatiron), Mike O'Neil (NYU), Alexander Novara (NYU), Ben Wandelt (JHU), Kaze Wong (JHU), and the Blanton--Hogg group meeting at NYU for valuable discussions.
The Flatiron Institute is a division of the Simons Foundation.
\Hogg{CH: You need to acknowledge a grant number here, I believe.}
The authors also note that, while we believe the formula is (barely) novel, it is difficult to be sure, given the challenge of conducting a complete literature review for this ancient topic.\par}

@@

>>>>

% bibliography issues
\usepackage{aas_macros,natbib}
\bibliographystyle{aasjournal}
\setcitestyle{aysep={}}
\setlength{\bibsep}{0pt plus 0.3ex}

@@

\medskip
\paragraph{Abstract}
Any permutation-invariant function of data points $\vec{r}_i$
can be written in the form $\rho(\sum_i\phi(\vec{r}_i))$ for suitable functions $\rho$ and $\phi$.
This form---known in the machine-learning literature as Deep Sets---also generates a map--reduce algorithm.
The area of a triangle is a permutation-invariant function of the locations $\vec{r}_i$ of the three corners $1\leq i\leq 3$.
We find the formula for the area of a triangle that is explicitly in Deep Sets form.
This project was motivated by questions about the fundamental computational complexity of $n$-point statistics in cosmology; that said, no insights of any kind were gained from these results.

@@

A function $f(\vec{r}_1, \dots, \vec{r}_N)$ is invariant to permutation of its arguments if and only if it can be decomposed in the form
\begin{equation}
    f(\vec{r}_1, \dots, \vec{r}_N) = \rho \, \big( \sum_{i=1}^{N} \, \phi(\vec{r}_i) \big)~,\label{eq:DeepSets}
\end{equation}
for some functions $\rho$ and $\phi$, where $\phi$ is usually vector valued.
This result---often known by the name of the deep learning architecture derived from it, \emph{Deep Sets} (\citealt{Zaheer+17deepsets}; see Theorem~2)---is not at all obvious.
For example, even though Heron's formula \eqref{eq:Heron} is explicitly permutation-invariant, it is not actually in the form of \eqref{eq:DeepSets}. 
Further, despite the triangle area being permutation-invariant, there is no known (prior to this work) formula for the area of a triangle that is explicitly in Deep Sets form.

While this result was recently popularized as Deep Sets, it follows from the classical theory of invariant polynomials.
It has long been known that any permutation-invariant (or symmetric) polynomial of scalar data $f(\{r_i\})$ can be written as a polynomial of the power sums $\sum_i r_i^m$ \citep{Waring1782}.
Further, symmetric polynomials of any-dimensional data are decomposable into \emph{polarizations} of the scalar symmetric polynomials \citep{Weyl1939}.
Together, these imply that a permutation-invariant polynomial of vector data is a polynomial in the polarizations of the power sums; e.g., the power sums of the components of the vectors.
This result is \emph{universal}, in the sense that it is true for all symmetric polynomials.
It also matches onto the Deep Sets form, where the polarizations of the power sums are the components of the vector $\sum_i \phi(\vec{r}_i)$, which are combined in a polynomial $\rho$.

% Specifically, it is known that the permutation-invariant polynomials on a set of any-dimensional data are representable in terms of the polarizations of the permutation-invariant polynomials on a corresponding set of one-dimensional data \citep{Weyl1939}.
% Further, the permutation-invariant polynomials on a set of one-dimensional data are representable in terms of the power sums $\sum_{i=1}^{n} r_i^j$ \citep{Waring1782}.
% Note that this decomposition is \emph{universal}, in the sense that it is true for \emph{all} symmetric polynomials of any-dimensional data. 
% This also matches onto the Deep Sets form, as it states that a permutation-invariant polynomial of any-dimensional data $f(\{r_i\})$ can be decomposed into a polynomial ($\rho$) of polarizations of the power sums ($\sum_i \phi(\vec{r}_i)$).
% \CH{this is mostly plagiarized from BBS email; need to understand it better to massage it.} 
% \Hogg{Important to emphasize that there is a proof of universality, and it's old.}

@@

Further, and important to our own motivations, cosmological clustering statistics such as 
the correlation function \citep{Peebles1973},
the power spectrum \citep{Peebles1973},
the three-point correlation function \citep{PeeblesGroth1975},
the bispectrum \citep{FrySeldner1982},
and higher-order statistics \citep{Peebles1980book}
used for precision measurement in cosmology \citep[e.g.,][]{Eisenstein+2005BAO,Planck18PNG,Planck18Inflation,Cabass+2022}
are all permutation-invariant.
Calculating these statistics na\"ively involves loops over all $k$-tuplets of points, which is not only inconsistent with Deep Sets form, but also has a runtime that scales as $N^k$, where $N$ the number of points in the set.

@@

\CH{moved this to here from farther down}
The area of a triangle is not only invariant to permutation, but also to translation, rotation, and reflection.
Formulae \eqref{eq:sine} and \eqref{eq:Heron}, by virtue of referencing only side-lengths and angles, are explicitly invariant to these other symmetries.
While true, it is not obvious that our solution \eqref{eq:result} is so invariant.
An explicit proof of this is left to the reader.

@@

\secbreak
% We now consider two more applications of the Deep Sets form to Heron's formula and to the potential energy of a system of point masses.

% \CH{I want to cut this?}
% We stated above that Heron's formula \eqref{eq:Heron} is explicitly permutation-invariant, and yet it is not represented in Deep Sets form. Consider instead
% \begin{equation}
% \label{eq:HeronDeepSets}
%     \Delta = \frac{1}{4} \, \sqrt{
%         \left( a^2 + b^2 + c^2 \right)^2
%         - 2 \left( a^4 + b^4 + c^4 \right)
%     }~.
% \end{equation}
% It is simple to show that this is exactly equivalent to Heron's formula when expanded, but it is now in Deep Sets form.\footnote{
%     This is not novel. Heron's formula is a special case of Brahmagupta's formula, which gives the area $K$ of a convex cyclic quadrilateral in terms of its four side lengths $a, b, c, d$ \citep{coolidge1939}.
%     On Wikipedia, it has been written
%     \begin{equation}
%         K = \frac{1}{4} \, \sqrt{
%             \left( a^2 + b^2 + c^2 + d^2 \right)^2
%             + 8 \, a\,b\,c\,d
%             - 2 \left( a^4 + b^4 + c^4 + d^4 \right)
%         }~,
%     \end{equation}
%     which is exactly \eqref{eq:HeronDeepSets} if $d$ is zero.
%     This expression is in Deep Sets form, with the product $a \, b \, c \, d$ 
%     representable as $\exp( \sum_{x \in \{ a,b,c,d \}} \log x )$.
% }
% 
While the triangle area is in a form similar to some physics problems, the polynomial methods we employ to find the Deep Sets form for the triangle area \eqref{eq:result} are not so relevant for those problems.
More relevant instead is to consider a $\phi$ which Fourier transforms the input data.
In doing so, translation symmetry is encoded as well as permutation symmetry, since the Fourier basis $\exp (i\, \vec{k}\cdot\vec{x})$ is the eigenbasis of the translation operator.

For example, this can be used to produce a Deep Sets form for the gravitational potential energy of a system of point masses. Explicitly, consider a system of $N$ point particles with masses $m_i$ and positions $\vec{r}_i$.
The total energy is given in Deep Sets form by
\begin{align}
    \phi_{\vec{k}}(m_j, \vec{r}_j)
    &=m_j \, e^{-i \, \vec{k} \cdot \vec{r}_j}
    ~,
    \quad
    \sum_{i=1}^{N} \phi_{\vec{k}}(m_i, \vec{r}_i)
    = \mathcal{F}[\rho](\vec{k})~,
    \nonumber\\
    U \, \big( \sum_{i=1}^{N} \, \phi_{\vec{k}}(m_i, \vec{r}_i) \big)
    &= -\frac{G}{4 \pi^2} \,
        \int \d^3 r \,
        \mathcal{F}^{-1} \left[
            \mathcal{F}[\rho] \ast \left(
            k^{-2} \, \mathcal{F}[\rho] \right)
        \right]~.
    \label{eq:EnergyDeepSets}
\end{align}
A derivation is given in the \nameref{sec:appendix}.
In principle this requires $\phi$ to be evaluated at every possible wavenumber $\vec{k}$.
To avoid making this infinite-dimensional $\phi$, one can instead evaluate $\mathcal{F}[\rho]$ at a finite set of wavenumbers $\vec{k}$ and live with approximations to the integrals necessary for the convolution and inverse Fourier transform.
The approximation will improve as the size of $\phi$ grows.
Approaches akin to this Fourier approach can be used \citep[and are used; e.g.,][]{Portillo+2018,Philcox+2022ENCORE} to estimate or approximate $k$-point statistics in cosmology, which was our motivating question.

We can even use these methods to find another formula for the area $\Delta$ of a triangle.
\begin{align}
    \phi_{\vec{k}} (\vec{r}_j) &= e^{-i \, \vec{k} \cdot \vec{r}_j}~,
    \quad
    \mathcal{F}[\rho](\vec{k}) = \sum_{j=1}^{3} \phi_{\vec{k}}(\vec{r}_j)~,
    \nonumber\\
    \mathcal{P}(\vec{r}) &= (2\pi)^d \, \mathcal{F}^{-1} \left[ 
    \abs{ \mathcal{F}[\rho] }^2 \right]~,
    \quad
    F_m = \int_{\mathbb{R}^d} \d^d r \, \mathcal{P}(\vec{r}) \, \frac{\abs{\vec{r}}^m}{2}~,
    \nonumber\\
    \Delta
    &= \frac{1}{4} \, \sqrt{F_2^2 - 2 \, F_4}~.
\end{align}
A complete derivation is provided in the \nameref{sec:appendix}.
Briefly, $\mathcal{P}$ is the Patterson function of X-ray crystallography \citep{Patterson1934} which has 6 delta function peaks corresponding to the three displacement vectors and their negatives. 
Multiplying the Patterson function by a function of the magnitude $r \equiv \vec{r}$ and integrating over all space gives a sum of the values of the function evaluated at the three side lengths. 
Choosing this function properly allows us to write the power sums of the side lengths $F_m = a^m+b^m+c^m$ in this form.

To verify this formula, note that we can write it in terms of the side lengths as
\begin{equation}
\label{eq:HeronDeepSets}
    \Delta = \frac{1}{4} \, \sqrt{
        F_2^2
        - 2 \, F_4
    }
    = \frac{1}{4} \, \sqrt{
        \left( a^2 + b^2 + c^2 \right)^2
        - 2 \, \left( a^4 + b^4 + c^4 \right)
    }~.
\end{equation}
This formula is equivalent to Heron's formula \eqref{eq:Heron} by expansion.
Note that it is also in Deep Sets form as a function of the side lengths.

\CH{I don't like this ending ... thinking about other ways to conclude}

We leave the reader with the following questions. 
It is trivial to take the form of \eqref{eq:result} and evaluate the sums for more than three vectors; does the result have any meaning?
Is there a generalization of the triangle area \eqref{eq:result} and simplex volume \eqref{eq:simplex} results to a $d$-simplex?
Our formulae \eqref{eq:result} and \eqref{eq:simplex} only used invariant factors of degree $2$, with exponents that are larger than $2$. Does this generalize to higher dimension simplexes, and can it be shown that higher order factors aren't needed?
One can replace the sum in \eqref{eq:DeepSets} with any other permutation-invariant aggregator (e.g., max or product), though we do not know if there are any claims to universality in these cases. Does another area formula exist for these other aggregators?
% We have discussed the general value of a map function $\phi$ which Fourier transforms the points;\footnote{%
% \Hogg{The Deep Sets version of the Fourier transform is reminiscent of a positional encoding (CITE), similar to those used to preserve ordering information in (otherwise permutation invariant) transformers (CITE).}}
% does a formula like \eqref{eq:EnergyDeepSets} exist for the area of a triangle?

{\footnotesize\par\bigskip\noindent
This project would not have been possible without originating inspiration and discussions with Soledad Villar (JHU).
The authors thank
  Ben Blum-Smith (JHU),
  Jonathan Goodman (NYU),
  Wilson Gregory (JHU),
  Teresa Huang (Flatiron),
  Alexander Novara (NYU),
  Mike O'Neil (NYU),
  Oliver Philcox (Columbia),
  Ben Wandelt (JHU),
  Kaze Wong (JHU),
and the Blanton--Hogg group meeting at NYU for valuable discussions.
The Flatiron Institute is a division of the Simons Foundation.
CH is supported by the National Science Foundation Graduate Research Fellowship under Grant Number DGE-2234660.
The authors also note that, while we believe the formula is (barely) novel, it is difficult to be sure, given the challenge of conducting a complete literature review for this ancient topic.\par}

@@

\paragraph{Fourier triangle area}
Suppose our triangle is specified by three vertices $\vec{r}_1,\vec{r}_2,\vec{r}_3 \in \mathbb{R}^d$. We represent this distribution by the field
\begin{equation}
    \rho(\vec{r})
    = \sum_i \, \delta(\vec{r} - \vec{r}_i)~.
\end{equation}
Fourier transforming this field gives
\begin{equation}
    \tilde{\rho}(\vec{k})
    =
    e^{-i \, \vec{k} \cdot \vec{r}_1} +
    e^{-i \, \vec{k} \cdot \vec{r}_2} +
    e^{-i \, \vec{k} \cdot \vec{r}_3}~.
\end{equation}
Multiplying by its conjugate to find the magnitude squared, we find
\begin{equation}
    \abs{\tilde{\rho}(\vec{k})}^2
    =
    3 +
    e^{-i \, \vec{k} \cdot \vec{r}_{12}} +
    e^{+i \, \vec{k} \cdot \vec{r}_{12}} +
    e^{-i \, \vec{k} \cdot \vec{r}_{13}} +
    e^{+i \, \vec{k} \cdot \vec{r}_{13}} +
    e^{-i \, \vec{k} \cdot \vec{r}_{23}} +
    e^{+i \, \vec{k} \cdot \vec{r}_{23}}~,
\end{equation}
defining $\vec{r}_{ij} \equiv \vec{r}_i - \vec{r}_j$.
The inverse Fourier transform of this is
\begin{align}
    \mathcal{P}(\vec{r})
    &= (2\pi)^d \, \mathcal{F}^{-1} \left[ \abs{\tilde{\rho}}^2 \right]
    \nonumber\\ 
    &= 3 \,\delta(\vec{r})
    + \delta(\vec{r} - \vec{r}_{12})
    + \delta(\vec{r} + \vec{r}_{12})
    \nonumber\\
    &\quad + \delta(\vec{r} - \vec{r}_{13})
    + \delta(\vec{r} + \vec{r}_{13})
    + \delta(\vec{r} - \vec{r}_{23})
    + \delta(\vec{r} + \vec{r}_{23})~.
\end{align}
This is known as the Patterson function.

By multiplying the Patterson function $\mathcal{P}$ by a function of $r \equiv \abs{\vec{r}}$ and integrating over all space, we can extract sums of functions of the side lengths. Explicitly,
\begin{align}
    \int_{\mathbb{R}^d} \d^d r \, \mathcal{P}(\vec{r}) \, f(\abs{\vec{r}})
    &= 3 \, f(0) +
        2 \, f(r_{12}) +
        2 \, f(r_{13}) +
        2 \, f(r_{23}) \nonumber\\
    &= 3 \, f(0) +
        2 \, f(a) +
        2 \, f(b) +
        2 \, f(c)~.
\end{align}
In particular, we can define the $k^{\rm th}$ power sum of the side lengths
\begin{equation}
    F_k
    = \int_{\mathbb{R}^d} \d^d r \, \mathcal{P}(\vec{r}) \, \frac{\abs{\vec{r}}^k}{2}
    = a^k + b^k + c^k~,
\end{equation}
which allows us to write the triangle area as
\begin{equation}
    \Delta
    = \frac{1}{4} \, \sqrt{\left(a^2+b^2+c^2\right)^2 - 2\, \left(a^4+b^4+c^4\right)}
    = \frac{1}{4} \, \sqrt{F_2^2 - 2\, F_4}~.
\end{equation}
